---
title: "decision_tree"
output: html_document
date: "2025-10-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 5.3 Decision Tree Modelling and Interpretation
### Step 1. Load Cleaned Dataset (XY)
Load the cleaned dataset from XY.csv.
Confirm that it is already in wide format (each row = one species, each column = one trait or derived feature).
```{r}

library(dplyr)
set.seed(2025)

# 1. Load and check target ----
XY <- read.csv("XY.csv")
stopifnot("y" %in% names(XY))
XY$y <- as.factor(XY$y)

# 2. Keep numeric predictors only ----
num_cols <- setdiff(names(XY)[sapply(XY, is.numeric)], c("y", "taxon_name"))
X <- XY[, num_cols, drop = FALSE]

# 3. Clean columns ----
X <- X[, colSums(!is.na(X)) >= 2, drop = FALSE]                    # remove mostly NA
X <- X[, apply(X, 2, sd, na.rm = TRUE) > 0, drop = FALSE]          # remove zero variance
is_bin <- sapply(X, function(z) all(z %in% c(0, 1), na.rm = TRUE)) # remove rare binary
if (any(is_bin)) {
  bin <- X[, is_bin, drop = FALSE]
  keep <- colMeans(bin, na.rm = TRUE) %>% between(0.005, 0.995)
  X <- cbind(X[, !is_bin, drop = FALSE], bin[, keep, drop = FALSE])
}

# 4. Remove rows with NA ----
good <- complete.cases(X)
model_df <- cbind(X[good, , drop = FALSE], y = droplevels(XY$y[good]))

# 5. Save ----
save(model_df, file = "model_df_clean.RData")

```

XY is fully loaded and clean.

The target variable is y, with 8 growth-form categories.

There are 654 numeric features ready for training.

  
### Step 2. Target Variable and Feature Check (Post-EDA Verification)

```{r}
library(dplyr)
library(ggplot2)
library(forcats)

# 1. Load cleaned dataset 
load("model_df_clean.RData")

# 2. Target summary 
target_summary <- model_df %>%
  count(y, name = "n") %>%
  mutate(pct = round(100 * n / sum(n), 2))

print(target_summary)

ggplot(target_summary, aes(x = fct_reorder(y, n), y = n)) +
  geom_col(fill = "skyblue") +
  coord_flip() +
  labs(title = "Class Distribution of y", x = "Class", y = "Count") +
  theme_minimal()

# 3. Quick numeric summary 
num_cols <- setdiff(names(model_df), "y")
summary(model_df[, head(num_cols, 5)])

```

### Step 3 Train/Test Split and Model Training (10-fold CV tuning)


```{r}

library(caret)
library(rpart)
library(rpart.plot)
library(MLmetrics)   # for multiClassSummary

set.seed(2025)

### 1. Load data (from Step 1) ----
# should give you model_df (predictors + factor y)
load("model_df_clean.RData")

### 2. Train / test split (70/30, stratified on y)
idx <- createDataPartition(model_df$y, p = 0.7, list = FALSE)
train <- model_df[idx, ]
test  <- model_df[-idx, ]

# quick check of class balance
round(prop.table(table(train$y)), 3)
round(prop.table(table(test$y)), 3)

# save split for later evaluation / confusion matrix etc.
save(train, test, file = "train_test_split.RData")

### 3. CV setup
# 10-fold CV, report Accuracy / Kappa / MacroF1 etc.
ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = multiClassSummary,
  allowParallel = FALSE  # keep single-threaded for reproducibility
)

### 4. Hyperparameter grid to search
# You can shrink these if training is too slow
grid <- expand.grid(
  cp       = seq(0.001, 0.05, by = 0.01),
  maxdepth = c(3, 5, 7, 9),
  minsplit = c(5, 10, 20, 30)
)

### 5. Custom rpart model (cp + maxdepth + minsplit)
rpart_custom <- list(
  label = "rpart (cp, maxdepth, minsplit)",
  library = "rpart",
  type = "Classification",

  parameters = data.frame(
    parameter = c("cp", "maxdepth", "minsplit"),
    class     = c("numeric", "numeric", "numeric"),
    label     = c("Complexity Parameter", "Max Depth", "Min Split")
  ),

  grid = function(x, y, len = NULL, search = "grid") {
    expand.grid(
      cp       = seq(0.001, 0.05, by = 0.01),
      maxdepth = c(3, 5, 7, 9),
      minsplit = c(5, 10, 20, 30)
    )
  },

  fit = function(x, y, wts, param, lev, last, classProbs, ...) {
    df <- as.data.frame(x)
    df$.outcome <- y

    rpart(
      .outcome ~ .,
      data = df,
      method = "class",
      control = rpart.control(
        cp       = param$cp,
        maxdepth = param$maxdepth,
        minsplit = param$minsplit
      ),
      ...
    )
  },

  predict = function(modelFit, newdata, submodels = NULL) {
    newdata <- as.data.frame(newdata)
    predict(modelFit, newdata, type = "class")
  },

  prob = function(modelFit, newdata, submodels = NULL) {
    newdata <- as.data.frame(newdata)
    as.data.frame(predict(modelFit, newdata, type = "prob"))
  },

  levels = function(x) x$ylevels,
  sort   = function(x) x[order(x$cp), ]
)

### 6. Train with caret + 10-fold CV
#dt_tuned <- train(
#  y ~ .,
#  data = train,
#  method = rpart_custom,
#  trControl = ctrl,
#  tuneGrid = grid,
#  metric = "Accuracy"   # multiClassSummary still calculates MacroF1 etc.
#)

# best tuning result

load("dt_tuned.RData")

print(dt_tuned)
dt_tuned$bestTune

# plot CV results (Accuracy vs params)
plot(dt_tuned, main = "CV Accuracy (Decision Tree Grid Search)")

# save tuned model for later (evaluation on test set, plotting tree, etc.)
save(dt_tuned, file = "dt_tuned.RData")
```


### Step 4. Train–Test Split
Before evaluating the final model, we divide the cleaned dataset into training and testing subsets.
The model will be trained on the training set (≈ 70%) and validated on the testing set (≈ 30%) to measure its generalization performance.

We also use the best parameters found in Step 3
(cp = 0.001, maxdepth = 7, minsplit = 10) to retrain the optimized decision tree model.

This ensures that the tree used for evaluation is both cross-validated and pruned appropriately.
```{r}
# Step 4: Train final decision tree with best params
library(dplyr)
library(caret)
library(rpart)
library(rpart.plot)

set.seed(2025)

### 1. Load train/test split from Step 3 
load("train_test_split.RData")

### 2. Best hyperparameters from CV (Step 3)
best_cp        <- 0.001
best_maxdepth  <- 7
best_minsplit  <- 10

### 3. Fit final model on training data
final_tree <- rpart(
  y ~ .,
  data = train,
  method = "class",
  control = rpart.control(
    cp       = best_cp,
    maxdepth = best_maxdepth,
    minsplit = best_minsplit
  )
)

### 4. Plot the tree
rpart.plot(
  final_tree,
  main = "Final Decision Tree",
  extra = 104,          # show predicted class + probs
  under = TRUE,
  faclen = 0,
  tweak = 0.8,
  cex = 0.5,
  fallen.leaves = TRUE,
  box.palette = "Blues"
)

### 5. Save model + data for later evaluation/report
save(final_tree, train, test, file = "final_decision_tree.RData")

```

### Step 5. Decision Tree Model Training

This step evaluates the performance of the final decision tree model trained in Step 4.
The model is applied to the test dataset (≈30%), which has not been used in training,
to assess its generalization ability.

```{r}
# Step 5: Evaluate final decision tree
library(caret)
library(yardstick)
library(ggplot2)
library(dplyr)

set.seed(2025)

### 1. Load model + data
# gives you: final_tree, train, test
load("final_decision_tree.RData")

### 2. Predict on test set
pred_class <- predict(final_tree, newdata = test, type = "class")

### 3. Confusion matrix + core metrics
cm <- confusionMatrix(pred_class, test$y)
print(cm)

# macro-style metrics (average over classes)
acc      <- cm$overall["Accuracy"]
precision<- mean(cm$byClass[, "Precision"], na.rm = TRUE)
recall   <- mean(cm$byClass[, "Recall"],    na.rm = TRUE)
f1_macro <- mean(cm$byClass[, "F1"],        na.rm = TRUE)
bal_acc  <- mean(cm$byClass[, "Balanced Accuracy"], na.rm = TRUE)

data.frame(
  Accuracy = round(acc, 3),
  Precision = round(precision, 3),
  Recall = round(recall, 3),
  MacroF1 = round(f1_macro, 3),
  BalancedAcc = round(bal_acc, 3)
)

### 4. Plot confusion matrix heatmap
cm_df <- as.data.frame(cm$table)

ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white", size = 3) +
  scale_fill_gradient(low = "skyblue", high = "darkblue") +
  labs(
    title = "Confusion Matrix - Final Decision Tree",
    x = "Actual",
    y = "Predicted"
  ) +
  theme_minimal(base_size = 12)

### 5. 95% bootstrap CI for accuracy
# resample test set with replacement and recompute accuracy
boot_iters <- 1000
acc_boot <- numeric(boot_iters)

for (i in 1:boot_iters) {
  idx <- sample(seq_len(nrow(test)), replace = TRUE)
  pred_i <- predict(final_tree, newdata = test[idx, ], type = "class")
  acc_boot[i] <- mean(pred_i == test$y[idx])
}

acc_ci <- quantile(acc_boot, c(0.025, 0.975))
acc_ci

### 6. Save results for the report
save(cm, acc, precision, recall, f1_macro, bal_acc, acc_ci,
     file = "decision_tree_evaluation.RData")

```
